Distributional Trees for Circular Data

### 2019-02-04 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
further revision
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2427 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: d37dc5fa62a5fde539e2ca4d2744abd872207fd4

In each subgroup maximum likelihood estimators $\hat \mu$ and $\hat \kappa$ are obtained
by maximizing the corresponding log-likelihood $\ell(\mu, \kappa; y) = \log(f_\mathrm{vM}(y;\mu, \kappa))$. 
The model scores are calculated by  $s(y; \mu, \kappa) = (\partial_{\mu} \ell(\mu, \kappa; y),
\partial_{\kappa} \ell(\mu, \kappa; y))$. In a subgroup of size $n$, evaluating the scores
at the individual observations and parameter estimates $s(y_i; \hat{\mu}, \hat{\kappa})$
yields an $n \times 2$ matrix that can be employed as a kind of residual, capturing how well
a given observation conforms with the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$, respectively.
Hence MOB can test for parameter instabilities in the model by assessing whether the scores
change along with the available covariates. If so, by maximimizing a partitioned likelihood
the parameter instabilities are incorporated into the model. This procedure is repeated recursively
until there are no significant parameter instabilities or until a another stopping criterion
is met (e.g., concerning subgroup sizes or depth of the tree).

@@

\section{Discussion / Summary and Outlook}
This study shows that applying distributional trees on circular data provides easily 
interpretable models clearly illustrating the impact of given covariates and their possible 
interactions on the distribution of the dependent variable. Moreover, covariates are selected
automatically such that no prespecification of the model is required which can be a challenging 
task, particularly for a meteorlogical quantity such as wind direction with high variation and 
many possible influencing factors.

Based on the results it can be assumed that including covariates such as lagged observations of 
wind direction as linear regressors provides valuable information. 
However, in order to properly account not only for the circular properties of the response but 
also of regressors, additional adaptions of distributional trees are required. In particular, 
if the selected split variable is a circular variable, e.g. wind direction, this demands for finding
two split point defining a ``split interval'' instead of a single split point.
This is also supported by the results illustrated in Figure~\ref{fig:tree} as mutliple splits in 
the linearly employed regressor \textit{dd} (lagged wind direction) are required in order to 
cluster the training data in corresponding wind directions.

@@

Another possible extension of the method as presented is to build an ensemble of 
distributional trees to regularize and stabilize the model 
(see, e.g., distributional forests, Schlosser et.~al, 2018). 
%In that way, it should not have a 
%significant impact on the results whether circular covariates are included as linear regressors
%yielding a larger number of splits or as circular regressors with ``intervall splits''.

>>>>

In each subgroup maximum likelihood estimators $\hat \mu$ and $\hat \kappa$ are obtained
by maximizing the corresponding log-likelihood $\ell(\mu, \kappa; y) = \log(f_\mathrm{vM}(y;\mu, \kappa))$. 
The model scores are calculated by  $s(y; \mu, \kappa) = (\partial_{\mu} \ell(\mu, \kappa; y),
\partial_{\kappa} \ell(\mu, \kappa; y))$. In a subgroup of size $n$, evaluating the scores
at the individual observations and parameter estimates $s(y_i; \hat{\mu}, \hat{\kappa})$
yields an $n \times 2$ matrix that can be employed as a kind of residual, capturing how well
a given observation conforms with the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$, 
respectively.
Hence MOB can test for parameter instabilities in the model by assessing whether the scores
change along with the available covariates. If so, by maximimizing a partitioned likelihood
the parameter instabilities are incorporated into the model. This procedure is repeated recursively
until there are no significant parameter instabilities or until another stopping criterion
is met (e.g., concerning subgroup sizes or depth of the tree).

@@

\section{Discussion and Outlook}
This study shows that applying distributional trees on circular data provides easily 
interpretable models clearly illustrating the impact of given covariates and their possible 
interactions on the distribution of the dependent variable. 
Moreover, covariates are selected automatically such that no prespecification of the model is 
required which can be a challenging task, particularly for a meteorlogical quantity such as wind 
direction with high variation and many possible influencing factors.
%Another valuable feature is the capability of capturing also non-additive and abrupt changes, 
Additionally, abrupt changes are captured, e.g., when the impact of a covariate such as 
\textit{air pressure difference} determines either up- or down-valley wind directions.

A possible extension of the method as presented is to build an ensemble of 
distributional trees to regularize and stabilize the model 
(see, e.g., distributional forests, Schlosser et.~al, 2018). 
%In that way, it should not have a 
%significant impact on the results whether circular covariates are included as linear regressors
%yielding a larger number of splits or as circular regressors with ``intervall splits''.

Note that the circular variable $dd$ (lagged observations of wind) is employed as a linear
regressor. Therefore, several splits are performed in $dd$ in order to properly cluster the training 
data in corresponding wind directions (see Figure~\ref{fig:tree}). 
In this case searching for two split points defining a split interval on the circular scale instead 
of a single split point on the linear scale could reduce the complexity of the model and possibly 
improve the results.
%Alternatively, in the case of a circular covariate 
%the tree building algorithm could search for two split points defining a split interval instead of 
%a single split point. This would reduce the complexity of the model and possibly improve the results.
%Based on the results it can be assumed that including covariates such as lagged observations of 
%wind direction as linear regressors provides valuable information. 
%However, in order to properly account not only for the circular properties of the response but 
%also of regressors, additional adaptions of distributional trees are required. In particular, 
%if the selected split variable is a circular variable, e.g. wind direction, this demands for finding
%two split point defining a ``split interval'' instead of a single split point.
%This is also supported by the results illustrated in Figure~\ref{fig:tree} as mutliple splits in 
%the linearly employed regressor \textit{dd} (lagged wind direction) are required in order to 
%cluster the training data in corresponding wind directions.  
\textit{Computational Details:} The employed \textsf{R} packages are available on \textsf{R}-Forge at
https://R-Forge.R-project.org/projects/partykit/.

@@
